For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
sec2,3

xxx

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, considering the signature for natural
numbers
\[
  \Gamma := \big(\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N\big),
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

In this section, we only need to check that the signatures and term model
construction in \cite{Kaposi:2019:CQI:3302515.3290315} restrict to IITs,
i.e.\ that from a syntax for the theory of IIT signatures all finitary IITs are
constructible.

\begin{enumerate}
\item
  Finitary IIT signatures are obtained exactly by dropping equality constructors from
  finitary QIIT signatures. It follows that any model for the theory of QIIT signatures
  restricts to IITs, hence we inherit the categorical semantics of QIITs.
\item
  The term model construction also restricts in a straightforward way. For QIITs,
  term algebras are constructed by induction on the syntax of signatures, and
  then another induction constructs the eliminator (i.e.\ dependent induction
  principle). In both cases, simply dropping equality constructors from the
  induction yields restriction to IITs.
\end{enumerate}

However, we shall mention that the Agda formalisation for the current paper
cannot be directly plugged into the Agda code for
\cite{Kaposi:2019:CQI:3302515.3290315} (which includes most of the term model
construction). One reason is that the QIIT formalisation uses strict computation
rules (given by Agda rewrite rules) for induction over signatures, while here we show propositional
computation. This mismatch can be in principle remedied by noting that both
formalisations use UIP and function extensionality, hence we can switch between
strict and propositional equalities, via the known translations between
extensional and intensional type
theories \cite{hofmann95extensional,winterhalter2019eliminating}.

Also, the current paper proves initiality, i.e.\ unique recursion for
signatures, while \cite{Kaposi:2019:CQI:3302515.3290315} uses dependent
elimination. We expect that the two notions are equivalent. An extension of
\cite{Kaposi:2019:CQI:3302515.3290315} to large infinitary QIITs would entail
this equivalence, since that would cover the (large, infinitary) theory of IIT
signatures.

>>>>

\begin{theorem}
  If a model of type theory supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}
